---
title: "Predictions using the Weight Lifting Exercises Dataset"
autor: "M.Welt"
output: html_document
date: "2025-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
Based on a dataset provide by HAR <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har> we will try to train a predictive model to predict what mistake was comitted during exercise.

We’ll take the following steps:

 - Process the data, for use of this project
 - Explore the data, try to reduce features
 - Train model
 - Predict exercise error on test dataset
 
---

```{r installing, echo=FALSE}
if (!requireNamespace("caret", quietly = TRUE)) {
    install.packages("caret")
}
if (!requireNamespace("randomForest", quietly = TRUE)) {
    install.packages("randomForest")
}
if (!requireNamespace("doParallel", quietly = TRUE)) {
    install.packages("doParallel")
}
if (!requireNamespace("formattable", quietly = TRUE)) {
    install.packages("formattable")
}
if (!requireNamespace("caretEnsemble", quietly = TRUE)) {
    install.packages("caretEnsemble")
}
if (!requireNamespace("xgboost", quietly = TRUE)) {
    install.packages("xgboost")
}
if (!requireNamespace("gbm", quietly = TRUE)) {
    install.packages("gbm")
}
if (!requireNamespace("ranger", quietly = TRUE)) {
    install.packages("ranger")
}
if (!requireNamespace("glmnet", quietly = TRUE)) {
    install.packages("glmnet")
}
```

```{r libraries, message=FALSE}

library(caret)    # For machine learning workflow
library(randomForest)  # For Random Forest model
library(ggplot2)  # For visualization
library(corrplot) # For correlation analysis
library(dplyr)    # For data manipulation
library(doParallel)
library(formattable)
library(caretEnsemble)
library(xgboost)
library(gbm)
library(ranger)
library(glmnet)

set.seed(123)  # Ensure reproducibility
```

## Acquire the datasets.

Download files if they are not yet on the system.

```{r donwloading, echo=TRUE,message=TRUE}
downloadFile <- function(fileName) {
  destfile <- paste0("pml-",fileName, ".csv")
  if (!file.exists(destfile)) {
    download.file(
      paste0("https://d396qusza40orc.cloudfront.net/predmachlearn/", fileName, ".csv"),
      destfile = destfile,
      mode = "wb"
    )
  }
}
downloadFile("training")
downloadFile("testing")
```

## Cleaning data

We will remove columns that are either empty or unsuitable for making predictions.

```{r cleaning_data }
cleanData <- function(fileName) {
  rawdata <- read.csv(paste0("pml-",fileName,".csv"))

 
  # Remove reduntant columns
    # Identifiers & Timestamps:
      #Remove columns that are time-based, likely irrelevant
      rawdata <- rawdata[,-c(grep("timestamp", names(rawdata)))]
      # "X" (index column)
      # "user_name" (not useful for prediction)
      rawdata <- rawdata[, !(names(rawdata) %in% c("X","user_name"))]
      #"new_window" and "num_window" (used for session tracking)
      rawdata <- rawdata[, !(names(rawdata) %in% c("new_window","num_window"))]
      
    # Remove columns where many values are NA
      # Define the threshold (90% missing values)
      threshold <- 0.9 * nrow(rawdata)
      rawdata <- rawdata[, colSums(is.na(rawdata) | rawdata == "") < threshold]
  #Set factor to "classe"    
  rawdata$classe <- as.factor(rawdata$classe)
  return(rawdata)
}  
cleaned_data <- cleanData("training")
```  
We have only `r ncol(cleaned_data)-1` features left.

## Split into Training and Validation sets
```{r splitting} 
trainIndex <- createDataPartition(cleaned_data$classe, p = 0.7, list = FALSE)
traindata <- cleaned_data[trainIndex, ]
validation_data <- cleaned_data[-trainIndex, ]
```   


#### Check missing values
```{r check_missing_values, message=FALSE}

missing_values_count <- sum(is.na(traindata))

```
We have `r missing_values_count` missing values.

#### Check near zero values
```{r near_zero_values, message=FALSE}
nzv_vars <- nearZeroVar(traindata, saveMetrics = TRUE)
count_nzv <- sum(nzv_vars$nzv == TRUE)

```
We have `r count_nzv` near zero values.

#### Remove Highly Correlated Features
Highly correlated features can lead to redundancy.
```{r remove_correlated, message=FALSE}

cor_matrix <- cor(traindata %>% select(-classe))  # Remove target variable
high_cor <- findCorrelation(cor_matrix, cutoff = 0.9)  # Identify high-correlation features
# traindata <- traindata[, -high_cor]

#Plot the correlation matrix
cor_matrix[abs(cor_matrix) < 0.6] <- NA  # Remove correlations < 0.5
cor_matrix[is.na(cor_matrix)] <- 0  # Replace NA with 0

corrplot(cor_matrix, method = "circle", type = "upper", 
         tl.cex = 0.7, tl.col = "black", 
         addCoef.col = NULL)  

traindata <- traindata[, -high_cor]

```
We have `r length(names(traindata))` columns left.

#### Parallel processing & Caching
Some processes are really time consuming, so let's prepare parallel processing
````{r parallel_functions}
# Corrected global cluster variable
parallelCluster <- NULL

# Function to stop the parallel cluster
stopParallelCluster <- function() {
  if (!is.null(parallelCluster)) {
    stopCluster(parallelCluster)  # Stop the parallel cluster
    registerDoSEQ()  # Reset to sequential execution
    parallelCluster <<- NULL  # Clear the global cluster variable
  }
}

# Function to prepare the parallel cluster
prepareParallelCluster <- function() {
  # Detect the number of available cores and use one less
  num_cores <- max(detectCores() - 1, 1)  # Ensure at least 1 core
  
  # Create and register the cluster globally
  parallelCluster <<- makeCluster(num_cores)
  registerDoParallel(parallelCluster)
}

# Function to load saved R objects from a .RData file if the file exists
loadObjects <- function(fileName) {
  # Append ".RData" extension to the filename
  fileName <- paste0(fileName, ".RData")
  
  # Check if the file exists before attempting to load
  if (file.exists(fileName)) {
    # Load objects into the global environment
    load(fileName, envir = .GlobalEnv)
  }
}
````

#### Train Multiple Base Models
Define a train_control object with cross-validation and set up your base models.

````{r train_models}
MODEL_LIST_FILE <- "model_list"
loadObjects(MODEL_LIST_FILE)
if (!exists("model_list") || is.null(model_list)) {
  start_time <- proc.time()
  prepareParallelCluster()
  # Define training control with 5-fold cross-validation
  train_control <- trainControl(
    method = "cv",
    number = 5,
    savePredictions = "final",
    allowParallel = TRUE
  )

  # List of base models
  model_list <- caretList(
    classe ~ .,  # Predicting Exercise Error
    data = traindata,
    trControl = train_control,
    methodList = c("ranger", "glmnet", "xgbTree")
  )
  
  stopParallelCluster()
  execution_time <- proc.time() - start_time
  save(model_list, execution_time, file = paste0(MODEL_LIST_FILE,".RData"))
} 
print(paste("Elapsed time:", round((execution_time["elapsed"] / 60), 2), "minutes"))
````
Explanation of the models:

- rf: Random Forest (good for non-linear data)
- gbm: Gradient Boosting Machine (great for medium datasets)
- xgbTree: XGBoost (optimized GBM)
- lm: Linear Regression (baseline model)

### Building the Stacked Model
Combine the base models using a meta-model (e.g., glm):
````{r}
STACK_MODEL_FILE <- "stacked_model"
loadObjects(STACK_MODEL_FILE)
if (!exists("stacked_model") || is.null(stacked_model)) {
  start_time <- proc.time()

  prepareParallelCluster()

  # Stack the models using a generalized linear model (GLM) as meta-learner
  stacked_model <- caretStack(
    model_list,
    method = "xgbTree",
    metric = "Accuracy",
    trControl = trainControl(
      method = "cv",
      number = 5,
      savePredictions = "final",
      classProbs = TRUE,  # Enable class probabilities for classification
      allowParallel = FALSE,
      sampling = "up"  # Upsample minority classes to avoid empty folds
    )
  )
  stopParallelCluster()
  execution_time <- proc.time() - start_time
  save(stacked_model, execution_time, file = paste0(STACK_MODEL_FILE,".RData"))
} 
print(paste("Elapsed time:", round((execution_time["elapsed"] / 60), 2), "minutes"))
````

